Geospatial - Assignment 3

### 2025-03-03 hw3/hw3_MoritzPeistNoemiLucchiSimonVellin.Rmd
Finalized submission 3: - notebook comments added - fixed isolation analysis

Next we plot the results and though the differences to our previous approach are minimal, the direct mean approach is more efficient and should be more what we are looking for:

@@

While both approaches yield similar results due to the linear nature of averaging, the direct regional calculation is methodologically preferred for spatial climate data analysis. In comparison the first graph to be replicated looked as follows:

@@

@@

# Part A

@@

- ne_10m_populated_places shapefile for the most populated cities 
- ne_10m_roads for the streets

@@

@@

## Methodological Approach

@@

for (i in 1:n_cities) {
  for (j in 1:n_cities) {
    if (i < j) {
      path <- shortestPath(tr_matrix,
        origin = as_Spatial(cities_planar[i, ]),
        goal = as_Spatial(cities_planar[j, ]),
        output = "SpatialLines"
      )
      path_sf <- st_as_sf(path)
      st_crs(path_sf) <- st_crs(cities_planar)
      # Calculate distance in kilometers
      dist <- as.numeric(st_length(path_sf)) / 1000
      dist_matrix[i, j] <- dist
      dist_matrix[j, i] <- dist
    }
  }
}

@@

## Plotting

@@

```{r}
# Plotting
ggplot() +
  geom_sf(data = st_transform(spain, 4326), fill = "lightgray") +
  geom_sf(data = st_transform(spain_roads, 4326), color = "black", size = 0.3) +
  geom_sf(data = st_transform(spain_top_10_cities, 4326), size = 1, color = "black") +
  geom_sf(
    data = st_transform(madrid_vigo %>% filter(NAME == "Madrid"), 4326),
    aes(shape = "Madrid", color = "Madrid"), size = 3
  ) +
  geom_sf(
    data = st_transform(madrid_vigo %>% filter(NAME == "Vigo"), 4326),
    aes(shape = "Vigo", color = "Vigo"), size = 3
  ) +
  scale_color_manual(
    name = "Cities",
    values = c("Madrid" = "red", "Vigo" = "blue")
  ) +
  scale_shape_manual(
    name = "Cities",
    values = c("Madrid" = 16, "Vigo" = 17)
  )
```

@@

  ![Map 1](data/ex2/raster07.png)
</div>

# Part B - Logical Approach

@@

```{r}
# Extract city names from the top 10 cities (excluding Madrid and Vigo for isolation analysis)
df_dist <- as.data.frame(dist_matrix) %>%
  rownames_to_column(var = "origin") %>%
  # Set up long format of the matrix
  gather(key = "destination", value = "distance", -origin) %>%
  # Exclude all other cities
  filter(origin %in% c("Madrid", "Vigo")) %>% 
  # Order first by origin, then by destination
  arrange(origin, destination)

@@

# Extract distances from Madrid to other cities
madrid_distances <- df_dist_wo_self %>%
  pull(distance)

# Extract distances from Vigo to other cities
vigo_distances <- df_dist_wo_self %>%
  pull(distance)
```

## Analysis

@@

## Plotting

To further explore the isolation of Madrid and Vigo, we create smoothed density distributions of their distances to the other top 10 cities. This visualization helps us understand the spread and concentration of distances, where a distribution with a wider spread or higher density at greater distances suggests greater isolation. We use geom_density() to plot the distributions, with Madrid in red and Vigo in turquoise, matching the style of the reference plot.

@@

In comparison the second graph looked as follows:

@@

# Part B - Alternative Approach

Although we think it seems logical to exclude Madrid and Virgo from the distance calculations, the reference plot seem to include them, as displayed by the noticeable density at x=0 for both cities. In this section, we computed the same plot but with "self-distances" included.

@@

>>>>

Next we plot the results and though the differences to our previous approach are minimal. However, the direct mean approach is more efficient and should be more what we are looking for:

@@

While both approaches yield similar results due to the linear nature of averaging, the direct regional calculation is methodologically preferred for spatial climate data analysis for several key reasons:

1. Spatial Weighting: The mean-of-means approach gives equal weight to each state regardless of its size, potentially overrepresenting smaller states in regional averages. The direct approach properly weights each grid cell based on its actual area contribution to the region.
2. Boundary Effects: State boundaries are political, not climatological. The mean-of-means approach artificially segments climate patterns along state lines before aggregation, while the direct approach preserves natural climate continuity across regions.
3. Statistical Accuracy: Multiple averaging steps in the mean-of-means approach can introduce compounding biases. The direct calculation minimizes this by reducing the number of aggregation steps.
4. Computational Efficiency: The direct approach requires fewer calculation steps and less memory, making it more efficient for large datasets.

While both methods yield similar results in this case (due to the linear nature of averaging and relatively uniform spatial distribution), the direct regional calculation better represents the true spatial characteristics of climate data.

In comparison the first graph to be replicated looked as follows:

@@

## Discussion of Plot Differences

After comparing both visualizations, the most likely reasons for the differences between our plot and the reference plot are:

- Different SPEI Timescale: Our code specifically uses spei01.nc (1-month SPEI), while the original plot likely uses a longer timescale index such as SPEI-12 or SPEI-24. Longer timescales produce smoother lines as they capture longer-term drought conditions rather than monthly fluctuations.
- Temporal Aggregation Method: Our code computes simple annual averages from monthly data, whereas the original plot might use a different aggregation approach (e.g., weighted seasonal averages or rolling averages) that reduces volatility.
- Smoothing Parameters: The original plot appears to use more aggressive smoothing for both regional lines and the trend line. Our geom_smooth() with default LOESS parameters may use different span or degree settings than the original.
- Data Pre-processing: The original may have applied additional data cleaning steps such as outlier removal or drought-specific thresholding that aren't present in our implementation.
- Spatial Aggregation Differences: The original plot might use different regional boundary definitions or apply population-weighted rather than area-weighted averaging.
- Year Range Subsetting: Despite both claiming 1965-2015, the exact start/end dates might differ slightly, affecting trend visualization, especially at the endpoints.
- Different SPEI Version: The SPEI calculation methodology itself has evolved over time, and different versions of the dataset might be used.

The most significant factor is likely the SPEI timescale difference, as this would fundamentally change the volatility patterns visible in the visualization.

@@

## Part A

### Data Loading

@@

- ne_10m_populated_places: for the most populated cities 
- ne_10m_roads: for the streets

@@

### Data Wrangling

@@

### Methodological Approach

@@

for (i in 1:n_cities) {
  for (j in 1:n_cities) {
    if (i < j) { # Only calculate the upper triangle and mirror
      path <- shortestPath(tr_matrix,
        origin = as_Spatial(cities_planar[i, ]),
        goal = as_Spatial(cities_planar[j, ]),
        output = "SpatialLines"
      )
      path_sf <- st_as_sf(path)
      st_crs(path_sf) <- st_crs(cities_planar)
      # Calculate distance in kilometers
      dist <- as.numeric(st_length(path_sf)) / 1000
      dist_matrix[i, j] <- dist
      # Mirror the distance in lower triangle
      dist_matrix[j, i] <- dist
    }
  }
}

@@

### Plotting - Map of Spain with Roads and Cities

@@

```{r}
# Plotting
ggplot() +
  geom_sf(data = st_transform(spain, 4326), fill = "lightgray") +
  geom_sf(data = st_transform(spain_roads, 4326), color = "black", size = 0.3) +
  geom_sf(data = st_transform(spain_top_10_cities, 4326), size = 1, color = "black") +
  geom_sf(
    data = st_transform(madrid_vigo %>% filter(NAME == "Madrid"), 4326),
    aes(shape = "Madrid", color = "Madrid"), size = 3
  ) +
  geom_sf(
    data = st_transform(madrid_vigo %>% filter(NAME == "Vigo"), 4326),
    aes(shape = "Vigo", color = "Vigo"), size = 3
  ) +
  scale_color_manual(
    name = "Cities",
    values = c("Madrid" = "red", "Vigo" = "blue")
  ) +
  scale_shape_manual(
    name = "Cities",
    values = c("Madrid" = 16, "Vigo" = 17)
  ) +
  theme(
    # Make background white
    panel.background = element_rect(fill = "white")
  )
```

@@

  ![Map 1](data/ex2/raster07.png)

## Part B - Distance Density Plots

### Data Wrangling

@@

```{r}
# Extract city names from the top 10 cities in long format
df_dist <- as.data.frame(dist_matrix) %>%
  rownames_to_column(var = "origin") %>%
  # Set up long format of the matrix
  gather(key = "destination", value = "distance", -origin) %>%
  # Exclude all other cities from origins
  filter(origin %in% c("Madrid", "Vigo")) %>% 
  # Order first by origin, then by destination
  arrange(origin, destination)

@@

# Extract distances from Madrid to other cities
madrid_distances <- df_dist_wo_self %>%
  filter(origin == "Madrid") %>%
  pull(distance)

# Extract distances from Vigo to other cities
vigo_distances <- df_dist_wo_self %>%
  filter(origin == "Vigo") %>%
  pull(distance)
```

### Analysis

@@

### Plotting - Distance Density Plot (excluding self-distances)

To further explore the isolation of Madrid and Vigo, we create smoothed density distributions of their distances to the other top 10 cities. This visualization helps us understand the spread and concentration of distances, where a distribution with a wider spread or higher density at greater distances suggests greater isolation. We use geom_density() to plot the distributions, with Madrid in red and Vigo in turquoise, matching the style of the reference plot. However, we exclude the self-distance of 0.

@@

In comparison the second reference graph looked as follows:

@@

### Plotting - Distance Density Plot (including self-distances)

Although we think it seems logical to exclude the self-distances of Madrid and Vigo from the distance calculations, the reference plot seems to include them, as displayed by the noticeable density at x=0 for both cities. Thus, in this section, we compute the same plot but with "self-distances" included.

@@

## Discussion of Plot Differences: Self-Distances and Calculation Parameters

### Self-Distances Methodological Considerations

The decision to include or exclude self-distances (city to itself) significantly impacts isolation analysis:

- **Excluding self-distances** (preferred for isolation studies) focuses on inter-city relationships, avoids artificial density spikes, and aligns with network analysis conventions.

- **Including self-distances** might be justified when measuring overall accessibility rather than isolation specifically, or when considering all possible origin-destination pairs in a transportation system.

### Geospatial Parameters Affecting Results

The differences between our plot and the reference are likely due to our choice of calculation parameters:

1. **Raster Resolution** (5000 units): Finer resolution would capture road networks more precisely but at computational cost.

2. **Friction Surface Values** (1:100 ratio): This strongly influences path selection and could dramatically change distances if the reference used different values.

3. **Road Network Data**: Differences in the Natural Earth road dataset completeness or classification.

4. **Transition Matrix Configuration**: The 8-direction setting affects possible path trajectories.

5. **Continental Spain Definition**: Different methods for isolating mainland Spain could affect which roads are included.

### Density Visualization Differences

Additional technical factors in the visualization itself:

1. **Bandwidth Parameter**: Different smoothing parameters in `geom_density()` would change curve shapes.

2. **Alpha Transparency and Color Settings**: Affect visual appearance of overlapping areas.

3. **Kernel Function**: Different mathematical functions for density estimation produce different curves.

Despite these differences, both plots support the same conclusion that Vigo is more isolated than Madrid, with Vigo's distribution showing consistently greater distances to other population centers.
